---
title: "Macular Degeneration Model"
author: "Corbin B, Matthew W"
format: revealjs
editor: visual
---

```{r, echo=FALSE, warning=FALSE}

library(patchwork)
library(tidyverse)
mac <- vroom::vroom("ARMD.txt")
mac$Trt <- as.factor(mac$Trt)

```

## 1. EDA - Basic Trends

```{r}

fill_trt <- scale_fill_manual(
  name = "Treatment",
  values = c("0" = "gray70", "1" = "steelblue"),
  labels = c("No", "Yes")
)

color_trt <- scale_color_manual(
  name = "Treatment",
  values = c("0" = "gray40", "1" = "steelblue"),
  labels = c("No", "Yes")
)

p1 <- mac %>%
  ggplot(aes(x = Vision,
             y = as.factor(Time),
             fill = as.factor(Trt))) +
  geom_violin(alpha = 0.8) +
  labs(
    title = "Distribution of Vision by Time and Treatment",
    x = "Vision Score",
    y = "Time (Weeks)",
    fill = "Treatment"
  ) +
  fill_trt+
  theme_minimal()


p2 <- mac %>%
  ggplot(aes(x = Baseline - Vision,
             y = as.factor(Time),
             fill = as.factor(Trt))) +
  geom_violin(alpha = 0.8) +
  labs(
    title = "Change in Vision from Baseline by Time",
    x = "Baseline − Vision (Change Score)",
    y = "Time (Weeks)",
    fill = "Treatment"
  ) +
  fill_trt+
  theme_minimal()


p3 <- mac %>%
  filter(Time == 52) %>%
  ggplot(aes(x = Vision, fill = as.factor(Trt))) +
  geom_histogram(
    bins = 20,
    alpha = 0.7,
    position = "identity"
  ) +
  labs(
    title = "Vision Distribution at Week 52",
    x = "Vision Score",
    y = "Count",
    fill = "Treatment"
  ) +
  fill_trt+
  theme_minimal()

p4 <- mac %>%
  ggplot(aes(x = Time,
             y = Vision,
             color = as.factor(Trt))) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    title = "Vision Over Time by Treatment",
    x = "Time (Weeks)",
    y = "Vision Score",
    color = "Treatment"
  ) +
  color_trt+
  theme_minimal()

treat_means <- mac %>%
  group_by(Time, Trt)%>%
  summarise(mean_vis = mean(Vision))

p5 <- treat_means %>%
  ggplot(aes(y=mean_vis, x = Time, color = Trt))+
  geom_line()+
  labs(
    title = "Interaction Plot: Mean Vision Score over time by Treatment",
    y = "Mean Vision Score Across Subjects"
  )+
  color_trt+
  theme_minimal()

p1+p3
```

## EDA Cont.

```{r}
p5
```

## EDA Cont.

```{r}

ind_lm <- lm(Vision ~ Time + Trt + Baseline, data = mac)

cor(matrix(residuals(ind_lm),ncol = 4, byrow = T))
```

This matrix helps demonstrate the correlation of a subjects vision over time. We see that there is a connection between a subject's different Vision Scores that should not be ignored. If we were to use a traditional linear regression model, this correlation would not be accounted for as part of the model.

## Model

We fit a **Generalized Least Squares (GLS)** model to account for the fact that each patient was measured four times — meaning observations within a subject are correlated and can't be treated as independent.

The model is:

> **Vision = β₀ + β₁(Baseline) + β₂(Time) + β₃(Trt) + β₄(Time × Trt) + ε**

## Why GLS

Each subject has 4 repeated measurements (weeks 4, 12, 24, 52). Measurements closer together in time tend to be more similar than measurements far apart. We used a compound symmetry / unstructured correlation structure (`corSymm`) to model this directly — this lets the data tell us how correlated any two timepoints are, rather than assuming they're all equally correlated or all independent.

## Justifying Assumptions (part 1)

Repeated measures correlation: Justified — patients were measured at the same 4 timepoints, and we'd expect a patient who sees well at week 4 to still see relatively well at week 12. Using corSymm lets us estimate this rather than assume it away.

Including Baseline as a covariate: Justified — patients entered the trial with very different vision scores (range: 31–85 letters). Without adjusting for baseline, we'd be comparing apples to oranges across groups.

## Justifying Assumptions (part 2)

Dropping Baseline × Trt interaction: Justified by the model selection F-test — the interaction was not significant, meaning the effect of baseline vision on follow-up vision does not differ between treatment groups.

Keeping Time × Trt interaction: Justified by the F-test — this interaction was significant, meaning the two groups decline at statistically different rates.

## Parameters in Context

Intercept (β₀) - Baseline Vision when all predictors = 0

Baseline (β₁) - Each additional letter of baseline vision predicts higher vision at follow-up — controls for where each patient started

Time (β₂) - The rate of vision loss per week for the control group

Treatment (β₃) - The average difference in vision between treated and control at Time = 0

Time × Treatment (β₄) - How much the treated group's slope differs from control — the treatment effect on the rate of decline

more junk

even more

## 3. Does interferon- stop vision loss over time? If not, how much does interferon-α slow

degeneration relative to no treatment?

## 4. Do patients on interferon-α have a higher vision score at 52 weeks

Yes they do!

On average, **we expect the vision score to be 8.241 points higher.**

Our 95% Confidence Interval is between 4.2415 and 12.2408.

## How we determined this

We calculated this using a generalized linear hypothesis test (glht) that lets us compare the expected difference of certain factors while holding all others constant.

In this case, our hypothesis was that :β₃ + β₄\*52 = 0

Keep in mind β₃ is the coefficient for the interferon-α treatment and β₄ is the coefficient for the interaction between the interferon-α treatment and time.

```{r}

library(nlme)

dat <- vroom::vroom("ARMD.txt")

# Make TAU (0) the baseline for Trt
dat <- dat %>%
  mutate(Trt = factor(Trt, levels = c(0, 1), labels = c("Control", "Treated")))


#FINAL MODEL

time_trt_mod <- gls(
  Vision ~ Baseline + Time * Trt,
  data = dat,
  correlation = corSymm(form = ~ 4 | Subject)
)



#On average and after controlling for the baseline score, do patients on interferon- have a
#significantly higher vision score at 52 weeks than those not on the treatment?


library(multcomp)

glht_52 <- glht(
  time_trt_mod,
  linfct = c(
    "TrtTreated + Time:TrtTreated * 52 = 0"
  )
)
library(multcomp)

glht_52 <- glht(
  time_trt_mod,
  linfct = c(
    "TrtTreated + Time:TrtTreated * 52 = 0"
  )
)

#summary(glht_52)
#confint(glht_52)
```

## Summary

-   Each patient was measured four times - observations within a subject are correlated

-   We modeled this correlation structure directly, and also the interaction between the treatment type and time on vision scores

-   Using this model we found

    -   Interferon-α does not stop vision loss, but it does slow it

    -   Patients on interferon-α have a significantly higher vision score at 52 weeks
